Simulating High-Dimensional Multivariate Data using the \pkg{bigsimr} Package

### 2021-05-25 000-introduction.Rmd
Revise for JSS submission up to Chapter 4

Massive high-dimensional (HD) data sets are now commonplace in many areas of scientific inquiry. As new methods are developed for data, a fundamental challenge lies in designing and conducting simulation studies to assess the operating characteristics of analyzing such proposed methodology --- such as false positive rates, statistical power, interval coverage, and robustness --- often with comparison to existing methods. Further, efficient simulation empowers statistical computing strategies, such as the parametric bootstrap [@Chernick2008] to simulate from a hypothesized null model, providing inference in analytically challenging settings. Such Monte Carlo (MC) techniques become difficult for high-dimensional data using existing algorithms and tools. This is particularly true when simulating massive multivariate, non-normal distributions, arising naturally in many fields of study.

As many have noted, it can be vexing to simulate dependent, non-normal/discrete data, even for low dimensional settings [@MB13; @XZ19]. For continuous non-normal multivariate data, the well-known NORmal To Anything (NORTA) algorithm [@Cario1997] and other copula approaches [@Nelsen2007] are well-studied, with flexible, robust software available [@Yan2007; @Chen2001]. Yet these approaches do not scale in a timely fashion to high-dimensional problems [@Li2019gpu]. For discrete data, early simulation strategies had major flaws, such as failing to obtain the full range of possible correlations (e.g., admitting only positive correlations: see @Park1996). While more recent approaches [@MB13; @Xia17; @BF17] have largely remedied this issue for low-dimensional problems, the existing tools are not designed to scale to high dimensions.

Another central issue lies in characterizing dependence between components in the high-dimensional random vector. The choice of correlation in practice usually relates to the eventual analytic goal and distributional assumptions of the data (e.g., non-normal, discrete, infinite support, etc). For normal data, the Pearson product-moment correlation describes the dependency perfectly. As we will see, however, simulating arbitrary random vectors that match a target Pearson correlation matrix is computationally intense [@Chen2001; @Xia17]. On the other hand, an analyst might consider use of nonparametric correlation measures to better characterize monotone, non-linear dependence, such as Spearman's $\rho$ and Kendall's $\tau$. Throughout, we focus on matching these nonparametric dependence measures, as our aim lies in modeling non-normal data and these rank-based measures possess invariance properties favorable in our proposed methodology. We do, however, provide Pearson matching with some caveats.

With all this in mind, we present a scalable, flexible multivariate simulation algorithm. The crux of the method lies in the construction of a Gaussian copula in the spirit of the NORTA procedure. Further, we introduce the `bigsimr` R package that provides high-performance software implementing our algorithm. The algorithm design leverages useful properties of nonparametric correlation measures, namely invariance under monotone transformation and well-known closed-form relationships between dependence measures for the multivariate normal (MVN) distribution.

Our study proceeds by providing background information, including a description of a motivating example application: RNA-sequencing (RNA-seq) breast cancer data. Then we describe and justify our simulation methodology and related algorithms. Next, we detail an illustrative low-dimensional example of basic use of the `bigsimr` R package. Then we proceed with Monte Carlo studies under various bivariate distributional assumptions to assess accuracy. We conclude the Monte Carlo evaluations by summarizing the computation time for increasingly higher dimensional vectors. After the MC evaluations, we simulate random vectors motivated by our RNA-seq example, evaluate the accuracy, and provide example statistical computing tasks, namely MC estimation of joint probabilities and evaluating HD correlation estimation efficiency. Finally, we discuss the method's utility, limitations, and future directions.

>>>>

Massive high-dimensional (HD) data sets are now commonplace in many areas of scientific inquiry. As new methods are developed for data analysis, a fundamental challenge lies in designing and conducting simulation studies to assess the operating characteristics of proposed methodology --- such as false positive rates, statistical power, interval coverage, and robustness --- often to compare to methods. Further, efficient simulation empowers statistical computing strategies, such as the parametric bootstrap [@Chernick2008] to simulate from a hypothesized null model, providing inference in analytically challenging settings. Such Monte Carlo (MC) techniques become difficult for HD dependent data using existing algorithms and tools. This is particularly true when simulating massive multivariate, non-normal distributions, arising in many fields of study.

As many have noted, it can be vexing to simulate dependent, non-normal/discrete data, even for low-dimensional (LD) settings [@MB13; @XZ19]. For continuous non-normal LD multivariate data, the well-known NORmal To Anything (NORTA) algorithm [@Cario1997] and other copula approaches [@Nelsen2007] are well-studied, with flexible, robust software available [@Yan2007; @Chen2001]. Yet these approaches do not scale in a timely fashion to HD problems [@Li2019gpu]. For discrete data, early simulation strategies had major flaws, such as failing to obtain the full range of possible correlations (e.g., admitting only positive correlations: see @Park1996). While more recent approaches [@MB13; @Xia17; @BF17] have largely remedied this issue for LD problems, the existing tools are not designed to scale to high dimensions for discrete marginal distributions.

Another central issue lies in characterizing dependence between components in the HD random vector. The choice of correlation in practice usually relates to the eventual analytic goal and distributional assumptions of the data (e.g., non-normal, discrete, infinite support, etc). For normal data, the Pearson product-moment correlation describes the dependency perfectly. As we will see, however, simulating arbitrary random vectors that match a target Pearson correlation matrix is computationally intense [@Chen2001; @Xia17]. On the other hand, an analyst might consider use of nonparametric correlation measures to better characterize monotone, non-linear dependence, such as Spearman's $\rho$ and Kendall's $\tau$. Throughout, we focus on matching these nonparametric dependence measures, as our aim lies in modeling non-normal data and these rank-based measures possess invariance properties favorable in our proposed methodology. We do, however, implement Pearson matching, but several layers of approximation is required.

With all this in mind, we present a scalable, flexible multivariate simulation algorithm. The crux of the method lies in the construction of a Gaussian copula in the spirit of the NORTA procedure. As we will describe in more detail, the algorithm's design leverages useful properties of nonparametric correlation measures, namely invariance under monotone transformation and well-known closed-form relationships between dependence measures for the multivariate normal (MVN) distribution. For our method, we developed a high-performance implementation: the `bigsimr` R package, written in native Julia code.

Our study proceeds by providing background information, including a description of a motivating example application: RNA-sequencing (RNA-seq) breast cancer data. Then we describe and justify our simulation methodology and related algorithms. We proceed by providing an illustrative LD `bigsimr` workflow. Next we conduct MC studies under various bivariate distributional assumptions to evaluate performance and accuracy. After the MC evaluations, we simulate random vectors motivated by our RNA-seq example, evaluate the accuracy, and provide example statistical computing tasks, namely MC estimation of joint probabilities and evaluating HD correlation estimation efficiency. Finally, we discuss the method's utility, limitations, and future directions.

### 2021-05-25 010-background.Rmd
Revise for JSS submission up to Chapter 4

The `bigsimr` `R` package presented here provides multiple algorithms that operate with high-dimensional multivariate data; however, all these algorithms were originally designed to support a single task: to generate random vectors drawn from multivariate probability distributions with given marginal distributions and dependency metrics. Specifically, our goal is to efficiently simulate a large number, $B$, of HD random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with *correlated* components and heterogeneous marginal distributions, described via cumulative distribution functions (CDFs) $F_i$.

@@

* BP1: A wide range of dependences, allowing both positive and negative values, and, ideally, admitting the full range of possible values.
* BP2: Flexible dependence, meaning that the number of bivariate marginals can be equal to the number of dependence parameters.
* BP3: Flexible marginal modeling, generating heterogeneous data --- possibly from differing probability families.

@@

Simulating high-dimensional, non-normal, correlated data motivates this work --- in pursuit of modeling RNA-sequencing (RNA-seq) data [@Wang2009b; @Conesa2016b] derived from breast cancer patients. The RNA-seq data-generating process involves counting how often a particular form of human messenger RNA (mRNA) is expressed in a biological sample. RNA-seq platforms typically quantify the entire transcriptome in one experimental run, resulting in high-dimensional data. For human-derived samples, this results in count data corresponding to over 20,000 genes (protein-coding genomic regions) or even over 77,000 isoforms when alternatively-spliced mRNA are counted [@Schissler2019]. Importantly, due to inherent biological processes, gene expression data exhibit correlation (co-expression) across genes [@BE07; @Schissler2018].

We illustrate our methodology using a well-studied Breast Invasive Carcinoma (BRCA) data set housed in The Cancer Genome Atlas (TCGA; see Acknowledgments).
For ease of modeling and simplicity of exposition, we only consider high expressing genes. In turn, we begin by filtering to retain the top `r d` of the highest-expressing genes (in terms of median expression) of the over 20,000 gene measurements from $N=`r nrow(example_brca)`$ patients' tumor samples. This gives a great number of pairwise dependencies among the marginals (specifically, $`r choose(d,2)`$ correlation parameters). Table \@ref(tab:ch010-realDataTab) displays RNA-seq counts for three selected high-expressing genes for the first five patients' breast tumor samples. To help visualize the bivariate relationships for these three selected genes across all patients, Figure \@ref(fig:ch010-realDataFig) displays the marginal distributions and estimated Spearman's correlations.

```{r ch010-realDataTab}
set.seed(2020-02-25)
num_genes    <- 3
num_patients <- 5
gene_sample  <- sample(example_genes[1:1000], num_genes)

@@

```{r ch010-realDataFig, cache=FALSE, fig.width=8, fig.cap="Marginal scatterplots, densities, and estimated pairwise Spearman's correlations for three example genes from Table 1. The data possess outliers, heavy-right tails, are discrete, and have non-trivial intergene correlations. Modeling these data motivate our simulation methodology."}
ggpairs(
  data = example_brca[, gene_sample],
  upper = list(continuous = wrap('cor', method = "spearman"))
) + 
  theme_bw() +
  theme(axis.text.x = element_text(angle = -90, vjust = 0.5))
```

@@

In multivariate analysis, an analyst must select a metric to quantify dependency.
The most widely-known is the Pearson (product-moment) correlation coefficient that describes the linear association between two random variables $X$ and $Y$, and, it is given by

@@

As @MB13 and @MK01 discuss, for a bivariate normal $(X,Y)$ random vector, the Pearson correlation completely describes the dependency between the components. For non-normal marginals with monotone correlation patterns, however, $\rho_P$ suffers some drawbacks and may mislead or fail to capture important relationships [@MK01]. Alternatively in these settings, analysts often prefer rank-based correlation measures to describe the degree of monotonic association.

Two nonparametric, rank-based measures common in practice are Spearman's correlation (denoted $\rho_S$) and Kendall's $\tau$. Spearman's $\rho_S$ has an appealing correspondence as the Pearson correlation coefficient on *ranks* of the values, thereby captures nonlinear yet monotone relationships. Kendall's $\tau$, on the other hand, is the difference in probabilities of concordant and discordant pairs of observations, $(X_i, Y_i)$ and $(X_j, Y_j)$, with concordance meaning that orderings have the same direction (e.g., if $X_i < X_j$, then $Y_i < Y_j$). Note that concordance is determined by the ranks of the values, not the values themselves.

@@

Given two marginal distributions, $\rho_P$ is not free to vary over the entire range of possible correlations $[-1,1]$. The so-called *Frechet-Hoeffding bounds* are well-studied [@Nelsen2007; @BF17]. These constraints cannot be overcome through algorithm design. In general, the bounds are given by

@@

There is a strong connection of our simulation strategy to Gaussian *copulas* (see @Nelsen2007 for a technical introduction). A copula is a distribution function on $[0,1]^d$ that describes a multivariate probability distribution with standard uniform marginals. This provides a powerful, natural way to characterize joint probability structures. Consequently, the study of copulas is an important and active area of statistical theory and practice.

@@

A Gaussian copula has marginal CDFs that are all standard normal, $F_i = \Phi, \forall \, i$. This representation corresponds to a multivariate normal (MVN) distribution with standard normal marginal distributions and covariance matrix ${\bf R_P}$. As the marginals are standardized to have unit variance, however, ${\bf R_P}$ is a Pearson correlation matrix. If $F_{{\bf R}}$ is the CDF of such a multivariate normal distribution, then the corresponding Gaussian copula $C_{{\bf R}}$ is defined through

@@

where $\Phi(\cdot)$ is the standard normal CDF. Note that the copula $C_{{\bf R}}$ is the familiar multivariate normal CDF of the random vector $(\Phi(X_1), \ldots, \Phi(X_d))$, where $(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_P})$.

@@

## Relevant Multivariate Simulation software Comparison

To place `Bigsimr`'s implementation and algorithms in context with existing approaches, as well as make this work's contribution more evident, we make some qualitative comparisons to other `R` multivariate simulation packages.  The `MASS` packages `mvnorm()` has long been available to produce random vectors from multivariate normal (and $t$) distributions. This familiar procedure cannot simulate from arbitrary margins and dependency measures, as well as fails to readily scale to high dimensions. Recently, the high-performance package `mvnfast` has been released that provides scaleable multivariate normal generation, but this doesn't completely satistify our goals of flexible margins and dependency modeling. For dependent discrete data, small-dimesional packages exist, such as `GenOrd`. The full-featured package `copula` provides many different copulas and a fairly general random generator `rCopula`. For low-dimensional modeling situtations, copula is suitable, yet the entent of exact matching of margins automatically is unclear. High-dimesional random vectors via `copula` takes an impractical amount of computing time [@Li2019gpu]. The `nortaRA` R package matches Pearson coefficients near exactly and is reasonable for low-dimenional settings. In sum, we know of no other `R` package that can produce HD random vectors with flexible margins and dependence modeling. Table (\#tab:compare-table) summarizes the above discussion.

\begin{table}[]
\centering
\caption{Selected feature comparison of `bigsimr` to other existing `R` multivariate simulation packages.}
\label{tab:compare-table}
\begin{tabular}{@{}llll@{}}
R Package & Scales to HD & Hetereogenous Margins & Flexible Dependence \\ \midrule
MASS      & No           & No                    & No                  \\
mvnfast   & Yes          & No                    & No                  \\
GenOrd    & No           & Discrete margins      & Yes                 \\
copula    & No           & Yes                   & No                  \\
nortaRA   & No           & Yes                   & No                  \\
bigsimr   & Yes          & Yes                   & Yes                
\end{tabular}
\end{table}

>>>>

The `bigsimr` `R` package presented here provides multiple algorithms that operate with HD multivariate data; however, all these algorithms were originally designed to support a single task: to generate random vectors drawn from multivariate probability distributions with given marginal distributions and dependency metrics. Specifically, our goal is to efficiently simulate a large number, $B$, of HD random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with *correlated* components and heterogeneous marginal distributions, described via cumulative distribution functions (CDFs) $F_i, i=1,\ldots,d$.

@@

* BP1: A wide range of dependences, allowing both positive and negative values, and, ideally, admitting the full range of possible values.
* BP2: Flexible dependence, meaning that the number of bivariate marginals can be equal to the number of dependence parameters.
* BP3: Flexible marginal modeling, generating heterogeneous data --- including mixed continuous and discrete margins.

@@

Simulating HD, non-normal, correlated data motivates this work, in pursuit of modeling RNA-sequencing (RNA-seq) data [@Wang2009b; @Conesa2016b] derived from breast cancer patients. The RNA-seq data-generating process involves counting how often a particular form of messenger RNA (mRNA) is expressed in a biological sample. RNA-seq platforms typically quantify the entire transcriptome in one experimental run, resulting in HD data. For human-derived samples, this results in count data corresponding to over 20,000 genes (protein-coding genomic regions) or even over 77,000 isoforms when alternatively-spliced mRNA are counted [@Schissler2019]. Importantly, due to inherent biological processes, gene expression data exhibit correlation (co-expression) across genes [@BE07; @Schissler2018].

We illustrate our methodology using the Breast Invasive Carcinoma (BRCA) data set housed in The Cancer Genome Atlas (TCGA; see Acknowledgments). For ease of modeling and simplicity of exposition, we only consider high expressing genes. In turn, we begin by filtering to retain the top `r d` of the highest-expressing genes (in terms of median expression) of the over 20,000 gene measurements from $N=`r nrow(example_brca)`$ patients' tumor samples. This gives a great number of pairwise dependencies among the marginals (specifically, $`r choose(d,2)`$ correlation parameters). Table \@ref(tab:ch010-realDataTab) displays RNA-seq counts for three selected high-expressing genes for the first five patients' breast tumor samples. To help visualize the bivariate relationships for these three selected genes across all patients, Figure \@ref(fig:ch010-realDataFig) displays the marginal distributions and estimated Spearman's correlations.

```{r tab:ch010-realDataTab}
set.seed(2020-02-25)
num_genes    <- 3
num_patients <- 5
gene_sample  <- sample(example_genes[1:1000], num_genes)

@@

```{r ch010-realDataFig, cache=FALSE, fig.width=8, fig.cap="Pairwise scatterplots for three example genes (Table 1), with estimated Spearman's correlations and marginal density estimates displayed. The data possess outliers, heavy-right tails, are discrete, and have non-trivial intergene correlations. Modeling these data motivate our simulation methodology."}
ggpairs(
  data = example_brca[, gene_sample],
  upper = list(continuous = wrap('cor', method = "spearman"))
) + 
  theme_bw() +
  theme(axis.text.x = element_text(angle = -90, vjust = 0.5))
```

@@

In multivariate analysis, an analyst must select a metric to quantify dependency. The most widely-known is the Pearson correlation coefficient that describes the linear association between two random variables $X$ and $Y$, and, it is given by

@@

As @MB13 and @MK01 discuss, for a bivariate normal $(X,Y)$ random vector, the Pearson correlation completely describes the dependency between the components. For non-normal marginals with monotone correlation patterns, however, $\rho_P$ suffers some drawbacks and may mislead or fail to capture important relationships [@MK01]. Alternatively, analysts often prefer rank-based correlation measures to describe the degree of monotonic association.

Two nonparametric, rank-based measures common in practice are Spearman's correlation (denoted $\rho_S$) and Kendall's $\tau$. Spearman's $\rho_S$ has an appealing correspondence as the Pearson correlation coefficient on *ranks* of the values, thereby capturing nonlinear yet monotone relationships. Kendall's $\tau$, on the other hand, is the difference in probabilities of concordant and discordant pairs of observations, $(X_i, Y_i)$ and $(X_j, Y_j)$ (concordance meaning that orderings have the same direction, e.g., if $X_i < X_j$, then $Y_i < Y_j$). Note that concordance is determined by the ranks of the values, not the values themselves.

@@

Given two marginal distributions, $\rho_P$ is not free to vary over the entire range of possible correlations $[-1,1]$. The *Frechet-Hoeffding bounds* are well-studied [@Nelsen2007; @BF17]. These constraints cannot be overcome through algorithm design. In general, the bounds are given by

@@

There is a connection of our simulation strategy to Gaussian *copulas* [@Nelsen2007]. A copula is a distribution function on $[0,1]^d$ that describes a multivariate probability distribution with standard uniform marginals. This provides a powerful, natural way to characterize joint probability. Consequently, the study of copulas is an important and active area of statistical theory and practice.

@@

A Gaussian copula has marginal CDFs that are all standard normal, $F_i = \Phi, \forall \, i$, where $\Phi(\cdot)$ is the standard normal CDF. This representation corresponds to a multivariate normal (MVN) distribution with standard normal marginal distributions and covariance matrix ${\bf R_P}$. Since the marginals are standardized to have unit variance, however, ${\bf R_P}$ is a Pearson correlation matrix. If $F_{{\bf R}}$ is the CDF of such a multivariate normal distribution, then the corresponding Gaussian copula $C_{{\bf R}}$ is defined through

@@

Here the copula function $C_{{\bf R}}$ is the familiar multivariate normal CDF of the random vector $(\Phi(X_1), \ldots, \Phi(X_d))$, where $(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_P})$.

@@

## Other Multivariate Simulation R Packages

To place `bigsimr`'s implementation and algorithms in context with existing approaches, here we qualitatively compare to other `R` multivariate simulation packages.  The `MASS` packages `mvnorm()` has long been available to produce random vectors from multivariate normal (and $t$) distributions. This familiar procedure cannot simulate from arbitrary margins and dependency measures, as well as fails to readily scale to high dimensions. Recently, the high-performance package `mvnfast` [@Fasiolo2016] has been released that provides HD multivariate normal generation, but this lacks flexibility in marginal models and dependency modeling. For dependent discrete data, LD packages exist, such as `GenOrd`. For copula computation, the full-featured package `copula` provides many different copulas and a fairly general LD random generator `rCopula()`. High-dimesional random vectors via `copula` takes an impractical amount of computing time [@Li2019gpu]. The `nortaRA` R package matches Pearson coefficients near exactly and is reasonable for LD settings. Finally, we know of no other `R` package that can produce HD random vectors with flexible margins and dependence modeling. Table (\#tab:compare-table) summarizes the above discussion.

\begin{table}[]
\centering
\caption{Selected feature comparison of bigsimr to existing R packages for multivariate simulation.}
\label{tab:compare-table}
\begin{tabular}{@{}llll@{}}
R Package & Scales to HD & Hetereogenous Margins & Flexible Dependence \\ \midrule
MASS      & No           & No                    & No                  \\
mvnfast   & Yes          & No                    & No                  \\
GenOrd    & No           & Discrete margins      & Yes                 \\
copula    & No           & Yes                   & No                  \\
nortaRA   & No           & Yes                   & No                  \\
bigsimr   & Yes          & Yes                   & Yes                
\end{tabular}
\end{table}
